---
title: "Preliminary_results_geographic_variation_plumage"
author: "Ore, MJ."
date: "2024-06-18"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, error = FALSE, warning = FALSE, message = FALSE)
```


```{r prep data - geographic variation in color, echo = FALSE}

library(tidyverse)
library(here)

here::i_am("results/Preliminary_results_geographic_variation_plumage.Rmd")

avg_img <- read.csv(here("data/BTBW_whole_specimen_Image_Analysis_measurements_averaged_allpop.csv"))

## select only the back 
dorsum <- avg_img %>% 
  filter(pl_code %in% c("d"))

# remove individual who's luminance mean is an order of magnitude greater than others
d_sub <- dorsum %>%
  filter(ID != "612918" )

#make population names
d_sub$pop <-paste(d_sub$County, d_sub$State.Province, sep = "_")

## order df by latitude
d_sub <- arrange(d_sub, lat)
# Reorder 'pop' based on the order of 'lat'
d_sub$pop <- factor(d_sub$pop, levels = unique(d_sub$pop))


```

# How does brightness of the dorsum vary across populations?

```{r brightness by pop, echo=FALSE}


ggplot(data = d_sub, aes(y = dblMean, x = pop)) + 
  geom_boxplot(width = 0.6, position = position_dodge(width = 0.7)) +
  labs(x = "Location", y = "Luminance (brightness)") +
  theme_classic() +
  theme(
    aspect.ratio = 0.75,
    text = element_text(size = 16),
    axis.text = element_text(size = 14)
  )

```
```{r brightness by indv, echo=FALSE}

z <- lm(dblMean~lat, data = d_sub, na.action = na.exclude)
summary(z)

ggplot(data = d_sub, aes(x = lat, y = dblMean)) + 
  geom_point(width = 0.6, position = position_dodge(width = 0.7)) +
  labs(x = "Location", y = "Luminance (brightness)") +
  geom_smooth(aes(x = lat, y = dblMean), method = "lm", col = "black")+
  theme_classic() +
  theme(
    aspect.ratio = 0.75,
    text = element_text(size = 16),
    axis.text = element_text(size = 14)
  )

```
##Short wavelength by pop

```{r sw by pop, echo=FALSE}

ggplot(data = d_sub, aes(y = swMean, x = pop)) + 
  geom_boxplot(width = 0.6, position = position_dodge(width = 0.7)) +
  labs(x = "Location", y = "Short Wavelength Mean") +
  theme_classic() +
  theme(
    aspect.ratio = 0.75,
    text = element_text(size = 16),
    axis.text = element_text(size = 14)
  )

```


# Correlations between body parts - Are dark backed birds darker overall? 

```{r prep df for correlation plots, echo = F}

library(viridis)

## select only brightness for ASY males
lum <- avg_img %>% 
  select(ID, pl_code, dblMean, dblSD, uvMean, uvSD, area, Age, pop, lat, lon) %>% 
  filter(ID != "612918" & Age == "ASY")

lum_summarized <- lum %>%
  group_by(ID, Age, pop, lat, lon, pl_code) %>%
  summarise(
    dblMean = mean(dblMean),
    dblSD = mean(dblSD),
    uvMean = mean(uvMean),
    uvSD = mean(uvSD),
    area = mean(area),
    .groups = 'drop'
  )

lum_wide <- lum_summarized %>% 
  pivot_wider(names_from = pl_code, values_from = c(dblMean, dblSD, uvMean, uvSD, area), names_sep = "_" )



```

## Dorsum vs throat

```{r d vs t plot, echo=FALSE}
# check for normality
hist(lum_wide$dblMean_d)
hist(lum_wide$dblMean_t)

shapiro.test(lum_wide$dblMean_d) # yes 
shapiro.test(lum_wide$dblMean_t) #no

# Perform Pearson's correlation test
c <- cor.test(lum_wide$dblMean_d, lum_wide$dblMean_t, method = "spearman")

# Extract the Pearson's cor (correlation coefficient) and p-value
rho <- c$estimate
p_value <- c$p.value

# Format the text to include rho and p-value, ready to be used in annotations
text_annotation <- sprintf("Spearman's Rho: %.2f, p-value: %.3f", rho, p_value)


ggplot(lum_wide)+
  geom_point(aes(x = dblMean_d, y = dblMean_t, col = lat))+
  scale_color_viridis_c(direction = -1)+
  geom_smooth(aes(x = dblMean_d, y = dblMean_t), method = "lm", col = "black")+
  labs(x = "Dorsum Brightness", y = "Throat Brightness")+
  annotate("text", x = Inf, y = Inf, label = text_annotation, 
           hjust = 1.1, vjust = 1.1, size = 5, color = "blue")+
  theme_classic()


```

## Dorsum vs wing covert

```{r d vs o plot, echo=FALSE}
# check for normality
hist(lum_wide$dblMean_d)
hist(lum_wide$dblMean_o)

shapiro.test(lum_wide$dblMean_d) #yes
shapiro.test(lum_wide$dblMean_o) #no

# Perform Pearson's correlation test
c <- cor.test(lum_wide$dblMean_d, lum_wide$dblMean_o, method = "spearman")

# Extract the Pearson's cor (correlation coefficient) and p-value
rho <- c$estimate
p_value <- c$p.value

# Format the text to include rho and p-value, ready to be used in annotations
text_annotation <- sprintf("Spearman's Rho: %.2f, p-value: %.3f", rho, p_value)


ggplot(lum_wide)+
  geom_point(aes(x = dblMean_d, y = dblMean_o, col = lat))+
  scale_color_viridis_c(direction = -1)+
  geom_smooth(aes(x = dblMean_d, y = dblMean_t), method = "lm", col = "black")+
  labs(x = "Dorsum Brightness", y = "Wing covert Brightness")+
  annotate("text", x = Inf, y = Inf, label = text_annotation, 
           hjust = 1.1, vjust = 1.1, size = 5, color = "blue")+
  theme_classic()


```



## Dorsum and belly

```{r d vs b plot , echo = F}
#check for normality
hist(lum_wide$dblMean_b)
shapiro.test(lum_wide$dblMean_b) #no

# Perform Pearson's correlation test
c <- cor.test(lum_wide$dblMean_d, lum_wide$dblMean_b, method = "spearman")

# Extract the Pearson's cor (correlation coefficient) and p-value
rho <- c$estimate
p_value <- c$p.value

# Format the text to include rho and p-value, ready to be used in annotations
text_annotation <- sprintf("Spearman's Rho: %.2f, p-value: %.3f", rho, p_value)

ggplot(lum_wide)+
  geom_point(aes(x = dblMean_d, y = dblMean_b, col = lat))+
  scale_color_viridis_c(direction = -1)+
  geom_smooth(aes(x = dblMean_d, y = dblMean_b), method = "lm", col = "black")+
  labs(x = "Dorsum Brightness", y = "Belly Brightness")+
  annotate("text", x = Inf, y = Inf, label = text_annotation, 
           hjust = 1.1, vjust = 1.1, size = 5, color = "blue")+
  theme_classic()


```


## Dorsum and wing spot - do darker birds have whiter wing spots?

```{r d and ws plot, echo = F}
#test for normality
hist(lum_wide$dblMean_w)
shapiro.test(lum_wide$dblMean_w) #yes

# Perform Pearson's correlation test
c <- cor.test(lum_wide$dblMean_d, lum_wide$dblMean_w, method = "pearson")

# Extract the Pearson's cor (correlation coefficient) and p-value
cor <- c$estimate
p_value <- c$p.value

# Format the text to include rho and p-value, ready to be used in annotations
text_annotation <- sprintf("Pearson's cor: %.2f, p-value: %.3f", cor, p_value)

ggplot(lum_wide)+
  geom_point(aes(x = dblMean_d, y = dblMean_w, col = lat))+
  scale_color_viridis_c(direction = -1)+
  geom_smooth(aes(x = dblMean_d, y = dblMean_w), method = "lm", col = "black")+
  labs(x = "Dorsum Brightness", y = "Wingspot Brightness")+
  annotate("text", x = Inf, y = Inf, label = text_annotation, 
           hjust = 1.1, vjust = 1.1, size = 5, color = "blue")+
  theme_classic()

```


# Do darker birds have larger wing spots?

```{r dorsum vs wingspot area, echo=FALSE}

shapiro.test(lum_wide$area_w) # yes
shapiro.test(lum_wide$dblMean_d) # yes


# Perform Pearson's correlation test
c <- cor.test(lum_wide$dblMean_d, lum_wide$area_w, method = "pearson")

# Extract the Pearson's cor (correlation coefficient) and p-value
cor <- c$estimate
p_value <- c$p.value

# Format the text to include rho and p-value, ready to be used in annotations
text_annotation <- sprintf("Pearson's cor: %.2f, p-value: %.3f", cor, p_value)


ggplot(lum_wide)+
  geom_point(aes(x = dblMean_d, y = area_w, col = lat))+
  scale_color_viridis_c(direction = -1)+
  geom_smooth(aes(x = dblMean_d, y = area_w), method = "lm", col = "black")+
  labs(x = "Dorsum Brightness", y = "Wingspot Area")+
  annotate("text", x = Inf, y = Inf, label = text_annotation, 
           hjust = 1.1, vjust = 1.1, size = 5, color = "blue")+
  theme_classic()


```

#Do brighter/larger wingspots correlate to throat badge or uv in crown?

```{r sex selection, echo=FALSE}

shapiro.test(lum_wide$area_w) # yes
shapiro.test(lum_wide$dblMean_w) # yes
shapiro.test(lum_wide$dblMean_t) # no
shapiro.test(lum_wide$uvMean_c) # yes

hist(lum_wide$area_w) 
hist(lum_wide$dblMean_w) 
hist(lum_wide$dblMean_t) 
hist(lum_wide$uvMean_c) 

# Perform Pearson's correlation test
c <- cor.test(lum_wide$dblMean_t, lum_wide$area_w, method = "spearman")

# Extract the Pearson's cor (correlation coefficient) and p-value
rho <- c$estimate
p_value <- c$p.value

# Format the text to include rho and p-value, ready to be used in annotations
text_annotation <- sprintf("Spearman's Rho: %.2f, p-value: %.3f", rho, p_value)


ggplot(lum_wide)+
  geom_point(aes(x = dblMean_t, y = area_w, col = lat))+
  scale_color_viridis_c(direction = -1)+
  geom_smooth(aes(x = dblMean_t, y = area_w), method = "lm", col = "black")+
  labs(x = "Throat Brightness", y = "Wingspot Area")+
  annotate("text", x = Inf, y = Inf, label = text_annotation, 
           hjust = 1.1, vjust = 1.1, size = 5, color = "blue")+
  theme_classic()

# Perform Pearson's correlation test
c <- cor.test(lum_wide$dblMean_t, lum_wide$dblMean_w, method = "spearman")

# Extract the Pearson's cor (correlation coefficient) and p-value
rho <- c$estimate
p_value <- c$p.value

# Format the text to include rho and p-value, ready to be used in annotations
text_annotation <- sprintf("Spearman's Rho: %.2f, p-value: %.3f", rho, p_value)


ggplot(lum_wide)+
  geom_point(aes(x = dblMean_t, y = dblMean_w, col = lat))+
  scale_color_viridis_c(direction = -1)+
  geom_smooth(aes(x = dblMean_t, y = dblMean_w), method = "lm", col = "black")+
  labs(x = "Throat Brightness", y = "Wingspot Brightness")+
  xlim(0.025,0.06)+
  annotate("text", x = Inf, y = Inf, label = text_annotation, 
           hjust = 1.1, vjust = 1.1, size = 5, color = "blue")+
  theme_classic()


# Perform Pearson's correlation test
c <- cor.test(lum_wide$dblMean_w, lum_wide$uvMean_c, method = "pearson")

# Extract the Pearson's cor (correlation coefficient) and p-value
cor <- c$estimate
p_value <- c$p.value

# Format the text to include rho and p-value, ready to be used in annotations
text_annotation <- sprintf("Pearson's cor: %.2f, p-value: %.3f", cor, p_value)


ggplot(lum_wide)+
  geom_point(aes(x = dblMean_w, y = uvMean_c, col = lat))+
  scale_color_viridis_c(direction = -1)+
  geom_smooth(aes(x = dblMean_w, y = uvMean_c), method = "lm", col = "black")+
  labs(x = "Wingspot Brightness", y = "UV of crown")+
  annotate("text", x = Inf, y = Inf, label = text_annotation, 
           hjust = 1.1, vjust = 1.1, size = 5, color = "blue")+
  theme_classic()


# Perform Pearson's correlation test
c <- cor.test(lum_wide$area_w, lum_wide$uvMean_c, method = "pearson")

# Extract the Pearson's cor (correlation coefficient) and p-value
cor <- c$estimate
p_value <- c$p.value

# Format the text to include rho and p-value, ready to be used in annotations
text_annotation <- sprintf("Pearson's cor: %.2f, p-value: %.3f", cor, p_value)


ggplot(lum_wide)+
  geom_point(aes(x = area_w, y = uvMean_c, col = lat))+
  scale_color_viridis_c(direction = -1)+
  geom_smooth(aes(x = area_w, y = uvMean_c), method = "lm", col = "black")+
  labs(x = "Wingspot Area", y = "UV of crown")+
  annotate("text", x = Inf, y = Inf, label = text_annotation, 
           hjust = 1.1, vjust = 1.1, size = 5, color = "blue")+
  theme_classic()


```
#Age and Markers

prep data
```{r prep df for age plots, echo = F}

library(viridis)

## select only brightness for ASY males
age <- avg_img %>% 
  select(ID, pl_code, dblMean, dblSD, uvMean, uvSD, area, Age, pop, lat, lon) %>% 
  filter(Age == "ASY" | Age == "SY" & ID != "612918")

age_summarized <- age %>%
  group_by(ID, Age, pop, lat, lon, pl_code) %>%
  summarise(
    dblMean = mean(dblMean),
    dblSD = mean(dblSD),
    uvMean = mean(uvMean),
    uvSD = mean(uvSD),
    area = mean(area),
    .groups = 'drop'
  )

age_wide <- age_summarized %>% 
  pivot_wider(names_from = pl_code, values_from = c(dblMean, dblSD, uvMean, uvSD, area), names_sep = "_" )


```

## Are wing spots brighter or larger in ASY vs SY males

```{r wingspot and age, echo=F}

t.test(dblMean_w ~ Age, data = age_wide)

ggplot(data = age_wide, aes(y = dblMean_w, x = Age)) + 
  geom_boxplot(width = 0.6, position = position_dodge(width = 0.7)) +
  labs(x = "Age", y = "Luminance of wingspot (brightness)") +
  theme_classic() +
  theme(
    aspect.ratio = 0.75,
    text = element_text(size = 16),
    axis.text = element_text(size = 14)
  )

t.test(area_w ~ Age, data = age_wide)


ggplot(data = age_wide, aes(y = area_w, x = Age)) + 
  geom_boxplot(width = 0.6, position = position_dodge(width = 0.7)) +
  labs(x = "Age", y = "Area of wingspot") +
  theme_classic() +
  theme(
    aspect.ratio = 0.75,
    text = element_text(size = 16),
    axis.text = element_text(size = 14)
  )

```

## Are backs darker in ASY vs SY males
```{r dorsum brightness vs age, echo = F}

t.test(dblMean_d ~ Age, data = age_wide)


ggplot(data = age_wide, aes(y = dblMean_d, x = Age)) + 
  geom_boxplot(width = 0.6, position = position_dodge(width = 0.7)) +
  labs(x = "Age", y = "Dorsum brightness") +
  theme_classic() +
  theme(
    aspect.ratio = 0.75,
    text = element_text(size = 16),
    axis.text = element_text(size = 14)
  )


```